---
title: "CgXBP1_analysis"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}

library(magrittr)
source("../R/chip_analysis_functions.R")
library(EnrichedHeatmap)
library(ggplot2)
library(patchwork)
```

# PolII ChIP-seq analysis

## Generate normalised gene expression values (Commonly known as FPKM, RPKM, TPM, etc. ) for PolII data.

To generate the normalised expression values for each gene aligned reads were processed by `MACS2` (Zhang et al., 2008), and `BigWig` files were generated using `bedSort` and `bedGraphToBigWig` commands from `UCSC Kent utils` (Kent et al., 2010). `Samtools` (version 1.9) was used to index the resultant `BAM` file and check for alignment statistics. Elongating RNAPII occupancy was then measured by `normalizeToMatrix` function from the R package `EnrichedHeatmap`. Normalized values then saved to .rds files (see R script `generate_normalised_gene_expression_matrix_from_polII_chip_data.R`) which are then used for downstream analysis and visualizations.

## PolII occupancy heatmap (WT)

```{r profile-hearmap-wt, message=FALSE, warning=FALSE, cache=TRUE}


gff_file <- "../data/C_glabrata_CBS138_version_s02-m07-r06_features.gff"
gff <- GenomicFeatures::makeTxDbFromGFF(gff_file, metadata = T)
genes <- GenomicFeatures::genes(gff)

## prepare signal data

bw_files_dir <- c("../data/bw/WT")

bw_files <- list.files(bw_files_dir, pattern = "*set1.*bw", recursive = T, full.names = T)

names(bw_files) <- c("0.5h", "2h", "4h", "6h", "8h")
bw_files <- tibble::enframe(bw_files)

## generate normalised matrix in tidy way
wt_norm_mat <- bw_files %>%
        dplyr::mutate(bw = purrr::map(value, function(ii) {
                rtracklayer::import(ii)
        })) %>%
        dplyr::mutate(norm_matrix = purrr::map(bw, function(ii) {
                nn <- EnrichedHeatmap::normalizeToMatrix(signal = ii, 
                                                         target = genes, 
                                                         value_column = "score", 
                                                         smooth = T, 
                                                         extend = 200)
                return(nn)
        }))

## plot hm list
wt_profile_heatmap <- get_enrichment_heatmap_list(
        x = wt_norm_mat$norm_matrix,
        col = circlize::colorRamp2(breaks = seq(0, 50, by = 10), colors = c("white", "#fdd49e", "#fdbb84", "#fc8d59", "#e34a33", "#b30000")),
        names = wt_norm_mat$name,
        titles = wt_norm_mat$name,
        axis_name = c("-200", "TSS", "TES", "+200"),
        axis_name_rot = 90 ,
        axis_name_gp = gpar(fontsize = 10, fontface = "bold"),
        pos_line = T,
        top_annotation = HeatmapAnnotation(lines = anno_enriched(axis_param = list(side = "right", facing = "inside")))
)

ComplexHeatmap::draw(wt_profile_heatmap)


```

## Prepare data for normalised gene expression heatmap

```{r prepare-data}

# load FPKM matrix 
wt_fpkm_mat <- readr::read_rds(file = "../data/rds/cg_polII_wt_thp1_sample_wise_fpkm.rds")   


wt_fpkm_mat <- wt_fpkm_mat %>%
        tidyr::gather("Cond", "FPKM", -gene_name) %>%
        tidyr::separate(col = "Cond", sep = "_", into = c("organism", "strain", "condition", "timepoint", "replicate"))


xbp1_fpkm_mat <- readr::read_rds(file = "../data/rds/cg_polII_xbp1_thp1_sample_wise_fpkm.rds")

xbp1_fpkm_mat <- xbp1_fpkm_mat %>%
        tidyr::gather("Cond", "FPKM", -gene_name) %>%
        tidyr::separate(col = "Cond", sep = "_", into = c("organism", "strain", "condition", "timepoint", "replicate"))


## combine datasets
cg_polII_data <- dplyr::bind_rows(wt_fpkm_mat, xbp1_fpkm_mat) %>%
        dplyr::group_by(gene_name, strain, condition, replicate) %>%
        dplyr::mutate(stdev = sd(FPKM)) ## calculate sd within group across all timepoints . Make sure that number of elem in each group must be equal to time points.

dplyr::bind_rows(wt_fpkm_mat, xbp1_fpkm_mat) %>% dplyr::group_by(gene_name, strain, condition, replicate) %>% 
        dplyr::tally() %>% 
        dplyr::arrange(desc(n))


## cross check the numbers (e.g number of unique genes, organism , strain, conditions etc.)
cg_polII_data %>% dplyr::ungroup() %>% dplyr::summarise_all(dplyr::n_distinct)


```

## Correlation of PolII replicates using FPKM values (WT)

```{r, fig.width=8, fig.height=6}


wt_data_for_corr <- cg_polII_data %>% dplyr::filter(strain == "wt") %>% 
  tidyr::pivot_wider(id_cols  = c(gene_name, timepoint), 
                     names_from = replicate, 
                     values_from = FPKM) 


plot_wt_cor <- wt_data_for_corr %>% 
  dplyr::mutate(set1 = log2(set1+1),set2 = log2(set2+1)) %>% 
  ggpubr::ggscatter(x = "set1",y = "set2", facet.by = "timepoint", cor.coef = T ,cor.coeff.args = list(col = "red"), nrow = 2 ) + 
  ggtitle("WT") + 
  scale_y_continuous(breaks = c(0,5,10)) + scale_x_continuous(breaks = c(0,5,10), limits = c(0,10))

plot_wt_cor
```

## Correlation of PolII replicates using FPKM values (xbp1)

```{r, fig.width=8, fig.height=6}
  
xbp1_data_for_corr <- cg_polII_data %>% dplyr::filter(strain == "xbp1") %>% 
  tidyr::pivot_wider(id_cols  = c(gene_name, timepoint), 
                     names_from = replicate, 
                     values_from = FPKM) 


plot_xbp1_cor <- xbp1_data_for_corr %>% 
  dplyr::mutate(set1 = log2(set1+1),set2 = log2(set2+1)) %>% 
  ggpubr::ggscatter(x = "set1",y = "set2", facet.by = "timepoint", cor.coef = T ,cor.coeff.args = list(col = "red"), nrow = 2 ) + ggtitle("xbp1∆")  +
  scale_y_continuous(breaks = c(0,5,10)) + scale_x_continuous(breaks = c(0,5,10), limits = c(0,10))


plot_xbp1_cor
```

## Plot SD distributions for each replicate

```{r sd-histograms}


sd_disturb_plot <- cg_polII_data %>%
        dplyr::slice(1L) %>% ## get one elem of each group
        ggplot() + geom_histogram(aes(x = stdev), binwidth = 1, fill = "cyan3", col = "black") + xlim(-1, 50) + facet_wrap(replicate ~ strain) + theme_bw()

sd_disturb_plot


```

## Count number of genes at several SD cutoffs (WT)

```{r count-genes-by-sd-wt, fig.width=4, fig.height=4}

gp <- counts_by_sd_cutoffs(
        x = wt_fpkm_mat %>%
                tidyr::spread(timepoint, FPKM) %>%
                dplyr::filter(replicate == "set1") %>%
                dplyr::select_if(is.double),
        y = paste(c(0, 2, 4, 6, 8), "h", sep = ""),
        sds_cutoffs = seq(1, 10, by = 0.5)
)

gp + theme_bw()
```

## PolII gene expression heatmap (WT)

```{r zscore-wt-heatmap, fig.width=3, fig.height=8}

## @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
####  zscore heatmap plot
## @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

sd_cut_off <- 2.25
strain_type <- "wt"
rep <- "set1"

fpkm_mat <- cg_polII_data %>%
        dplyr::ungroup() %>%
        dplyr::filter(stdev >= sd_cut_off, replicate == rep) %>%
        #dplyr::filter(strain == strain_type, stdev >= sd_cut_off, replicate == rep) %>%
        dplyr::select(gene_name, strain, timepoint, FPKM) %>%
        tidyr::unite(col = "ident", sep = "_", strain, timepoint, remove = F) %>%
        dplyr::select(-timepoint) %>%
        dplyr::group_by(gene_name)

zscore_mats <- fpkm_mat  %>%
        dplyr::group_by( strain, gene_name) %>% 
        dplyr::mutate(zscore = scale(FPKM)) %>% 
        dplyr::select(-FPKM) %>% 
        tidyr::spread(key = ident, value = zscore) %>%
        dplyr::group_by(strain) %>% 
        dplyr::group_split() %>%
        purrr::map(~ ..1 %>% dplyr::select(-strain) %>% TidyWrappers::tbl_remove_vars_NA_any())


# plot WT heatmap
hm_data <- zscore_mats[[1]] %>%
        as.data.frame() %>% 
        tibble::column_to_rownames("gene_name")


seed <- 7173
set.seed(seed)
wt_km <- kmeans(hm_data, centers = 8)
table(wt_km$cluster)
wt_cl_order <- factor(wt_km$cluster, levels = c(3,2,1,6,4,7,8,5)) %>% 
        forcats::fct_relabel(~paste("clust",1:8, sep = "-"))

cg_wt_heatmap = ComplexHeatmap::Heatmap(hm_data, show_row_dend = F, 
                        show_row_names = F, 
                        cluster_columns = F, 
                        show_column_dend = F, 
                        row_split = wt_cl_order, 
                        name = "Z-Score",
                        cluster_row_slices = F)
cg_wt_heatmap <- ComplexHeatmap::draw(cg_wt_heatmap)

```

## PolII gene expression heatmap (xbp1)

```{r zscore-xbp1-heatmap, fig.height=6, fig.width=3}

hm_data <- zscore_mats[[2]] %>%
        as.data.frame() %>% 
        tibble::column_to_rownames("gene_name")


seed <- 7173
set.seed(seed)
km <- kmeans(hm_data, centers = 7)
table(km$cluster)
cl_order <- factor(km$cluster, levels = c(4,6,1,5,3,2,7)) %>% 
        forcats::fct_relabel(~paste("clust",1:7, sep = "-"))

cg_xbp1_heatmap = ComplexHeatmap::Heatmap(hm_data, show_row_dend = F, 
                        show_row_names = F, 
                        cluster_columns = F, 
                        show_column_dend = F, 
                        row_split = cl_order, 
                        name = "Z-Score",
                        cluster_row_slices = F)

cg_xbp1_heatmap

```

## Plot transcription factor (WT) Temporally 

```{r, fig.width=4, fig.height=10}
cg_tf <- readr::read_lines("../data/Cg_polII_wt_temporal_TFs.txt")

tf_hm_data <- wt_cl_order[names(wt_cl_order) %in% cg_tf] %>% 
  tibble::enframe(name = "gene_name", value = "clust") %>% dplyr::left_join(zscore_mats[[1]] , by = "gene_name") %>% 
        as.data.frame() %>% 
        tibble::column_to_rownames("gene_name") 


cg_tf_heatmap = ComplexHeatmap::Heatmap(tf_hm_data[,-1], show_row_dend = F, 
                        show_row_names = T, 
                        row_names_gp = gpar(cex = 0.7),
                        cluster_columns = F, 
                        show_column_dend = F,
                        column_title_gp = gpar(cex = 0.8),
                        row_split = tf_hm_data$clust, 
                        name = "Z-Score",
                        column_title = "Temporally Expressed Transcription Factors",
                        cluster_row_slices = F)

ComplexHeatmap::draw(cg_tf_heatmap)
```

## Prepare data for pathway specific heatmap

```{r}

genes_by_processes <- tibble::tribble(
                                      ~gene_name,                ~processes,
                              "CAGL0A01045g",               "TCA Cycle",
                              "CAGL0D06424g",               "TCA Cycle",
                              "CAGL0E01287g",               "TCA Cycle",
                              "CAGL0F02431g",               "TCA Cycle",
                              "CAGL0G02673g",               "TCA Cycle",
                              "CAGL0G08712g",               "TCA Cycle",
                              "CAGL0H03993g",               "TCA Cycle",
                              "CAGL0L06798g",               "TCA Cycle",
                              "CAGL0A00363g", "Amino Acid Biosynthesis",
                              "CAGL0A02574g", "Amino Acid Biosynthesis",
                              "CAGL0C01243g", "Amino Acid Biosynthesis",
                              "CAGL0D04356g", "Amino Acid Biosynthesis",
                              "CAGL0F01749g", "Amino Acid Biosynthesis",
                              "CAGL0F06875g", "Amino Acid Biosynthesis",
                              "CAGL0G06732g", "Amino Acid Biosynthesis",
                              "CAGL0J06402g", "Amino Acid Biosynthesis",
                              "CAGL0E01243g",        "Iron Homeostasis",
                              "CAGL0G06798g",        "Iron Homeostasis",
                              "CAGL0H03487g",        "Iron Homeostasis",
                              "CAGL0J05060g",        "Iron Homeostasis",
                              "CAGL0L11990g",        "Iron Homeostasis",
                              "CAGL0M08910g",        "Iron Homeostasis",
                              "CAGL0L09251g",         "Stress Response",
                              "CAGL0G10175g",         "Stress Response",
                              "CAGL0G01540g",         "Stress Response",
                              "CAGL0L05412g",         "Stress Response"
                              )

id_map <- tibble::tribble(
                           ~gene_name,    ~gene_symbol,
                  "CAGL0A01045g",   "CgFUM1",
                  "CAGL0D06424g",   "CgACO1",
                  "CAGL0E01287g",   "CgKGD2",
                  "CAGL0F02431g",   "CgACO2",
                  "CAGL0G02673g",   "CgIDH1",
                  "CAGL0G08712g",   "CgKGD1",
                  "CAGL0H03993g",   "CgCIT1",
                  "CAGL0L06798g",   "CgMDH3",
                  "CAGL0A00363g",    "CgLEU",
                  "CAGL0A02574g",   "CgTRP4",
                  "CAGL0C01243g",   "CgHIS5",
                  "CAGL0D04356g",   "CgGCV1",
                  "CAGL0F01749g",   "CgSHM2",
                  "CAGL0F06875g",   "CgLYS1",
                  "CAGL0G06732g",   "CgLEU9",
                  "CAGL0J06402g",  "CgLYS20",
                  "CAGL0L09251g",   "CgHAL1",
                  "CAGL0G10175g",   "CgAWP6",
                  "CAGL0G01540g", "CgNCE103",
                  "CAGL0L05412g",   "CgASF1",
                  "CAGL0E01243g",   "CgCTH2",
                  "CAGL0G06798g",   "CgLSO1",
                  "CAGL0H03487g",   "CgAFT1",
                  "CAGL0J05060g",   "CgZAP1",
                  "CAGL0L11990g",   "CgGRX4",
                  "CAGL0M08910g",   "CgSNF1"
                  )


```

## Pathway specific heatmap: WT - TCA cycle

```{r fig.width=4, fig.height=5 }

data_for_tca_heatmap_wt <- genes_by_processes %>% 
  dplyr::left_join(id_map)  %>%
  dplyr::left_join(zscore_mats[[1]]) %>%
  dplyr::filter(processes == "TCA Cycle") %>%
  dplyr::select(-c(processes,gene_name)) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("gene_symbol")
  
  
ComplexHeatmap::Heatmap(data_for_tca_heatmap_wt, cluster_columns = F, show_row_dend = F, name = "Z-score", cluster_rows = F, column_title = "TCA Cycle")


```

## Pathway specific heatmap: xbp1 - TCA cycle

```{r fig.width=4, fig.height=5 }

data_for_tca_heatmap_xbp1 <- genes_by_processes %>% 
  dplyr::left_join(id_map)  %>%
  dplyr::left_join(zscore_mats[[2]]) %>%
  dplyr::filter(processes == "TCA Cycle") %>%
  dplyr::select(-c(processes,gene_name)) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("gene_symbol")


ComplexHeatmap::Heatmap(data_for_tca_heatmap_xbp1, cluster_columns = F, show_row_dend = F, name = "Z-score", cluster_rows = F, column_title = "TCA Cycle")



```

## Pathway specific heatmap: WT - Amino acid biosynthesis

```{r fig.width=4, fig.height=5 }

data_for_aa_heatmap_wt <- genes_by_processes %>% 
  dplyr::left_join(id_map)  %>%
  dplyr::left_join(zscore_mats[[1]]) %>%
  dplyr::filter(processes == "Amino Acid Biosynthesis") %>%
  dplyr::select(-c(processes,gene_name)) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("gene_symbol")
  
  
ComplexHeatmap::Heatmap(data_for_aa_heatmap_wt, cluster_columns = F, show_row_dend = F, name = "Z-score", cluster_rows = F, column_title = "Amino acid biosynthesis")


```

## Pathway specific heatmap: xbp1 - Amino acid biosynthesis

```{r fig.width=4, fig.height=5 }

data_for_aa_heatmap_xbp1 <- genes_by_processes %>% 
  dplyr::left_join(id_map)  %>%
  dplyr::left_join(zscore_mats[[2]]) %>%
  dplyr::filter(processes == "Amino Acid Biosynthesis") %>%
  dplyr::select(-c(processes,gene_name)) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("gene_symbol")


ComplexHeatmap::Heatmap(data_for_aa_heatmap_xbp1, cluster_columns = F, show_row_dend = F, name = "Z-score", cluster_rows = F, column_title = "Amino acid biosynthesis")


```
